State of the Practice for Lattice Boltzmann Method Software

### 2022-10-08 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updates to paper
Update paper to the plots and discussion for the actual measured data (not the sensitivity analysis data)

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of 20 questions). After measuring, we rank the software using the Analytic
    Hierarchy Process (AHP). Our ranking is roughly consistent with GitHub stars
    ranking, suggesting at least a partial correlation between the use of best
    practices and popularity. We find the state of the practice to be healthy,
    with 87.5\% of the measured packages ranking in the top five for at least one
    quality, the majority of LBM generated artifacts corresponding to general
    recommendations from research software developers, common use of version
    control (67\% of packages) and the adoption of a quasi-agile development
    process.  Areas of best practice to potentially improve include adoption of
    continuous integration, API documentation and enforcement of programming
    style guides.  We interviewed four developers to gain insight into their
    current pain points. Identified challenges include lack of development time,
    lack of funding, and difficulty with ensuring correctness.  Developers are
    addressing these pain points by designing for change, circumventing the
    oracle problem and prioritizing documentation and usability. For future
    improvements we suggest the following: employing linters, conducting
    rigorous peer reviews, writing and submitting more papers on software,
    growing the number of contributors by following current recommendations for
    open source projects, and augmenting the theory manuals to include more
    requirements specification relevant information.

@@

We investigate the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}. The specific
application of the methodology to LBM is reviewed in Section~\ref{methodology},
with the full details in Anonymous (xxxx)\footnotemark[1].
%\citet{Michalski2021}.  
Our methodology updates
an approach used in prior assessments of domains like Geographic Information
Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, statistical software for psychology
\citep{SmithEtAl2018_StatSoft} and medical image analysis software
\citep{Dong2021}.

@@

Once we measure each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP),
a decision-making technique that compares multiple options by multiple criteria.
Our AHP algorithm uses the overall impression score, gathered via the
measurement template, along with pairwise comparisons between packages to rank
the LBM software. We have previously used this technique to rank scientific
software \citep{SmithEtAl2016}.

@@

Figure~\ref{Fig_Installability} shows the installability ranking of the software
packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
DL\_MESO, Palabos) tend to have one set of linear installation instructions,
written as if the person doing the installation has none of the dependencies
installed. The instructions often list compatible operating system versions and
include instructions for installing dependencies. The top-ranked packages often
incorporate some automation of the installation process. The number of
dependencies a package has does not correlate with a higher score. The ability
to validate the installation process, say through tutorials or test cases, leads
to a higher score. Apparently active work on a project improves installability
since the top six ranked packages are all classified as alive.

@@

Figure~\ref{Fig_CorrectnessVerifiability} shows the ranking for surface
correctness and verifiability. Software packages with a higher score tend to
have theory documentation. They also explicitly use at least one document
generation tool that helps with verification, thus building confidence in
correctness. The top ranked software packages all include getting started
tutorials, and most of these include expected output. The two top ranked
packages, Ludwig and Musubi, indicate the use of unit testing in their
documentation. They also incorporated CI in the development process.
Furthermore, all of the top 5 ranked packages are alive.

@@

Figure~\ref{Fig_VisibilityTransparency} shows the visibility and transparency
ranking of the software packages. Software packages with a high score have an
explicit development model and defined development process. They also have
detailed and easy to access software release notes. Five of the top six ranked
packages are alive.

@@

Figure~\ref{Fig_OverallScore} shows the overall ranking. The AHP method finds
the weighting of quality $i$ ($w_i$) in the same way that the score ($s_j$) was
found for software package $j$ evaluated for a given quality
(Section~\ref{AHP}).  That is, we conduct a pairwise comparison between the
priority of different qualities to construct the $m \times m$ matrices $A$ and
$B$ and then we average row $i$ of matrix $B$ to find the priority value $w_i$
for quality $i$.  If we introduce the notation that $s^i_j$ is the score for
quality $i$ for package $j$, then the overall score $S_j$ for package $j$ is
found via:

@@

Notable exceptions to this trend occur for three projects that rank high by star
count, but relatively low in the ``follow best practices'' ranking: TCLB (star
rank 3, our rank 10), lettuce (star rank 4, our rank 12) and \esp{} (star rank
5, our rank 15).  Part of the discrepancy is because our process ranks all
projects, but the star measure does not apply to all projects.  Also, our
process equally weights all qualities (Section~\ref{Sec_OverallQuality}), but
for real users some qualities will be more important than others. Discrepancies
between our measures and the star measures may also be due to inaccuracy with
using stars to approximate popularity, because of how people use stars, because
young projects have less time to accumulate stars, and because stars represent
the community's feeling in the past more than current preferences
\citep{Szulik2017}. Older packages (like \esp{} and TCLB) have had more time to
accumulate stars and watches. Young projects have less time to accumulate stars,
like HemeLB, which only has 12 stars (at the time of measurement) having
recently moved from a different GitHub repository.  Lettuce is a young project
(released in 2019) with a relatively high number of stars. The discrepancy in
rankings may be due to a mismatch between the time we manually measured lettuce
and when we counted the stars. The late addition of Musubi means a gap between
when we measured projects and the time when our automated data (like stars) was
collected. A final reason for inconsistencies between our ranking and the
community's ranking is that, as for consumer products, more factors influence
popularity than just quality.

@@

Based on interviews with four developers, this section aims to answer the
research questions: i) What are the pain points for developers working on
research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
points of developers from LBM compare to the pain points for research software
in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
identified pain points and include citations that contrast the LBM experience
with observations from researchers in other domains.
Section~\ref{Sec_AddressConcerns} covers potential ways to address the pain
points and Anonymous (xxxx)\footnotemark[1] %\citet{SmithEtAl2021} 
provides the full interview questions.

@@

Current LBM software mostly follows best practices, as illustrated by
Table~\ref{topperformerstable}, which summarizes the top five performers for
each quality and for the overall quality. Taking the union of all entries in the
table, we have 21 top performers: ESPREsSo, MechSys, OpenLB, DL\_MESO, Palabos, Sailfish, Ludwig, Musubi, pyLBM, waLBerla, laboetie, lettuce, LUMA, TCLB, \esp, LB2D-Prime, LB3D-Prime, LIMBES, SunlightLB, HemeLB, MP-LABS.  Therefore, seven eights (87.5\%) of the 24 packages
appear in the top five for at least one quality.  Most entries are alive.  Alive packages are in
a healthy state with respect to adopting best practices.

\begin{table}[ht!]
	\begin{center}
		\begin{tabular}{ p{3cm}p{13cm} }
			\toprule
			Quality & Ranked 1st or 2nd\\
			\midrule
			Installability & ESPResSo, MechSys, OpenLB, DL\_MESO (LBE),
			Palabos, Sailfish\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig,
			Musubi, pyLBM\\%, ESPResSo, waLBerla\\
			\addlinespace[0.4cm]
			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*, lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB\\
			\addlinespace[0.4cm]
			Surface Robustness & DL\_MESO, ESPResSo, \esp, laboetie*, LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys, OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla, Musubi\\
			\addlinespace[0.4cm]
			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish, DL\_MESO, \esp LB2D-Prime*, LB3D, LUMA, TCLB\\
			\addlinespace[0.4cm]
			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla, \esp, LUMA, MP-LABS, pyLBM, TCLB\\
			\addlinespace[0.4cm]
			Reusability & ESPResSo, waLBerla, \esp, HemeLB*, Musubi, OpenLB, Palabos, Sailfish\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Surface \\Understandability} & LUMA, DL\_MESO (LBE), lettuce, Musubi, Palabos\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig, ESPResSo, LUMA, Palabos\\
			\addlinespace[0.4cm]
			Overall Quality & ESPResSo, Ludwig\\%, Palabos, LUMA, OpenLB\\
			\bottomrule
		\end{tabular}
		\caption{Top performers for each quality (all packages were alive at the
		time of measurement, except for those marked by an asterisk (*)) (Sorted
		by Order of Quality Measurement)}
		\label{topperformerstable}
	\end{center}
\end{table}

>>>>

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 108 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of 20 questions). After measuring, we rank the software using the Analytic
    Hierarchy Process (AHP). Our ranking is roughly consistent with GitHub stars
    ranking, suggesting at least a partial correlation between the use of best
    practices and popularity. We find the state of the practice to be healthy,
    with 67\% of the measured packages ranking in the top two for at least two
    qualities, the majority of LBM generated artifacts corresponding to general
    recommendations from research software developers, common use of version
    control (67\% of packages) and the adoption of a quasi-agile development
    process.  Areas of best practice to potentially improve include adoption of
    continuous integration, API documentation and enforcement of programming
    style guides.  We interviewed four developers to gain insight into their
    current pain points. Identified challenges include lack of development time,
    lack of funding, and difficulty with ensuring correctness.  Developers are
    addressing these pain points by designing for change, circumventing the
    oracle problem and prioritizing documentation and usability. For future
    improvements we suggest the following: employing linters, conducting
    rigorous peer reviews, writing and submitting more papers on software,
    growing the number of contributors by following current recommendations for
    open source projects, and augmenting the theory manuals to include more
    requirements specification relevant information.

@@

We investigate the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021, SmithAndMichalski2022}. The specific
application of the methodology to LBM is reviewed in Section~\ref{methodology},
with the full details in \citet{Michalski2021}.  
Our methodology updates
an approach used in prior assessments of domains like Geographic Information
Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, statistical software for psychology
\citep{SmithEtAl2018_StatSoft} and medical image analysis software
\citep{Dong2021}.

@@

Once we measure each package, we still need to rank them to answer
\rqref{RQ_HighestQuality}. We do this via the Analytic Hierarchy Process (AHP)
\citep{Saaty1980}, a decision-making technique that compares multiple options by
multiple criteria. Our AHP algorithm uses the overall impression score, gathered
via the measurement template, along with pairwise comparisons between packages
to rank the LBM software. We have previously used this technique to rank
scientific software \citep{SmithEtAl2016}.

@@

Figure~\ref{Fig_Installability} shows the installability ranking of the software
packages. Software packages with a higher score (ESPResSo, MechSys, OpenLB,
DL\_MESO, Palabos, Sailfish) tend to have one set of linear installation
instructions, written as if the person doing the installation has none of the
dependencies installed. The instructions often list compatible operating system
versions and include instructions for installing dependencies. The top-ranked
packages often incorporate some automation of the installation process. The
number of dependencies a package has does not correlate with a higher score. The
ability to validate the installation process, say through tutorials or test
cases, leads to a higher score. Apparently active work on a project improves
installability since the top six ranked packages are all classified as alive.

@@

Figure~\ref{Fig_CorrectnessVerifiability} shows the ranking for surface
correctness and verifiability. Software packages with a higher score tend to
have theory documentation. They also explicitly use at least one document
generation tool that helps with verification, thus building confidence in
correctness. The top ranked software packages all include getting started
tutorials, and most of these include expected output. Two of the top ranked
packages, Ludwig and Musubi, indicate the use of unit testing in their
documentation. They also incorporated CI in the development process.
Furthermore, all the top 5 ranked packages are alive.

@@

Figure~\ref{Fig_VisibilityTransparency} shows the visibility and transparency
ranking of the software packages. Software packages with a high score have an
explicit development model and defined development process. They also have
detailed and easy to access software release notes. All the top five ranked
packages are alive.

@@

Figure~\ref{Fig_OverallScore} shows the overall ranking. The AHP method finds
the weighting of quality $i$ ($w_i$) in the same way that the score ($s_j$) was
found for software package $j$ evaluated for a given quality
(Section~\ref{AHP}).  That is, we conduct a pairwise comparison between the
priority of different qualities to construct the $m \times m$ matrices $A$ and
$B$, and then we average row $i$ of matrix $B$ to find the priority value $w_i$
for quality $i$.  If we introduce the notation that $s^i_j$ is the score for
quality $i$ for package $j$, then the overall score $S_j$ for package $j$ is
found via:

@@

Notable exceptions to this trend occur for three projects that rank high by star
count, but relatively low in the ``follow best practices'' ranking: TCLB (star
rank 3, our rank 10), lettuce (star rank 4, our rank 12) and \esp{} (star rank
5, our rank 14).  Part of the discrepancy is because our process ranks all
projects, but the star measure does not apply to all projects.  Also, our
process equally weights all qualities (Section~\ref{Sec_OverallQuality}), but
for real users some qualities will be more important than others. Discrepancies
between our measures and the star measures may also be due to inaccuracy with
using stars to approximate popularity, because of how people use stars, because
young projects have less time to accumulate stars, and because stars represent
the community's feeling in the past more than current preferences
\citep{Szulik2017}. Older packages (like \esp{} and TCLB) have had more time to
accumulate stars and watches. Young projects have less time to accumulate stars,
like HemeLB, which only has 12 stars (at the time of measurement) having
recently moved from a different GitHub repository.  Lettuce is a young project
(released in 2019) with a relatively high number of stars. The discrepancy in
rankings may be due to a mismatch between the time we manually measured lettuce
and when we counted the stars. The late addition of Musubi means a gap between
when we measured projects and the time when our automated data (like stars) was
collected. A final reason for inconsistencies between our ranking and the
community's ranking is that, as for consumer products, more factors influence
popularity than just quality.

@@

Based on interviews with four developers, this section aims to answer the
research questions: i) What are the pain points for developers working on
research software projects (\rqref{RQ_PainPoints})?; and, ii) How do the pain
points of developers from LBM compare to the pain points for research software
in general (\rqref{RQ_ComparePainPoints})?  Below we go through each of the
identified pain points and include citations that contrast the LBM experience
with observations from researchers in other domains.
Section~\ref{Sec_AddressConcerns} covers potential ways to address the pain
points and \citet{SmithEtAl2021} provides the full interview questions.

@@

Current LBM software mostly follows best practices, as illustrated by
Table~\ref{topperformerstable}, which summarizes the top two packages for each
quality and for the overall quality. Taking the union of all entries in the
table that appear under 2 or more qualities, we have 16 top performers:
ESPREsSo, MechSys, OpenLB, DL\_MESO (LBE), Sailfish, Ludwig, Musubi, pyLBM,
waLBerla, laboetie, lettuce, \esp, LUMA, Palabos, TCLB and LB2D-Prime.
Therefore, two thirds (67\%) of the 24 packages appear in the top two for at
least two qualities.  Almost all the top performers are alive, except for two
packages: laboetie and LB2D-Prime.  Alive packages are in a healthy state with
respect to adopting best practices.

\begin{table}[ht!]
	\begin{center}
		\begin{tabular}{ p{3cm}p{13cm} }
			\toprule
			Quality & Ranked 1st or 2nd\\
			\midrule
			Installability & ESPResSo, MechSys, OpenLB, DL\_MESO (LBE),
			Palabos, Sailfish\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Surface Correctness \\ and Verifiability} & Ludwig,
			Musubi, pyLBM\\
			\addlinespace[0.4cm]
			Surface Reliability & Musubi, DL\_MESO (LBE), ESPResSo, laboetie*,
			lettuce, LUMA, MechSys, OpenLB, Palabos, pyLBM, TCLB\\
			\addlinespace[0.4cm]
			Surface Robustness & DL\_MESO (LBE), ESPResSo, \esp, laboetie*,
			LB2D-Prime*, LB3D-Prime*, lettuce, LIMBES*, Ludwig, LUMA, MechSys,
			OpenLB, Palabos, pyLBM, Sailfish, SunlightLB*, TCLB, waLBerla,
			Musubi\\
			\addlinespace[0.4cm]
			Surface Usability & ESPResSo, laboetie*, Ludwig, OpenLB, Sailfish,
			DL\_MESO (LBE), \esp, LB2D-Prime*, LB3D*, LUMA, TCLB\\
			\addlinespace[0.4cm]
			Maintainability & ESPResSo, Ludwig, Palabos, waLBerla, \esp, LUMA,
			MP-LABS*, pyLBM, TCLB\\
			\addlinespace[0.4cm]
			Reusability & ESPResSo, waLBerla, \esp, HemeLB*, Musubi, OpenLB,
			Palabos, Sailfish\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Surface \\Understandability} & LUMA, DL\_MESO (LBE),
			lettuce, Musubi, Palabos\\
			\addlinespace[0.4cm]
			\pbox{3.0cm}{Visibility and \\Transparency} & Ludwig, ESPResSo,
			LUMA, Palabos\\
			\addlinespace[0.4cm]
			Overall Quality & ESPResSo, Ludwig\\%, Palabos, LUMA, OpenLB\\
			\bottomrule
		\end{tabular}
		\caption{Top performers for each quality (all packages were alive at the
		time of measurement, except for those marked by an asterisk (*)) (Sorted
		by Order of Quality Measurement)} \label{topperformerstable}
	\end{center}
\end{table}
